---
title: "A Practical Guide to Descriptive Analysis in R"
subtitle: "Understanding your data before you model it"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    number-sections: true
execute:
  warning: false
  message: false
---

## Introduction

Every good data analysis begins in the same place: not with a machine learning model, not with a p-value, and not with a fancy visualization, but with a quiet, careful look at the data itself. This process is called **descriptive analysis** (or descriptive statistics), and it is one of the most undervalued skills in the analytics toolkit. It's the difference between an analyst who understands their data and one who is simply running code against it.

Descriptive analysis answers a deceptively simple question: *what does this dataset actually look like?* How many rows and columns does it have? What are the typical values? How spread out are they? Are there outliers, missing values, or strange distributions lurking beneath the surface? Long before you fit a regression, train a classifier, or run a hypothesis test, you need to answer these questions. Skipping this step is how analysts end up building models on data riddled with data-entry errors, skewed distributions, or hidden subgroups that quietly bias every result downstream.

R is, arguably, the best environment in existence for this kind of work. It was built by statisticians, for statisticians, and descriptive analysis is baked into its DNA. A single function call — `summary()` — will give you more useful information about a dataset than many other languages will give you in ten lines of code. Layer on packages like **dplyr**, **skimr**, **psych**, and **ggplot2**, and you have a complete, expressive toolkit for exploring any dataset that lands on your desk.

This article walks through descriptive analysis in R from first principles to more advanced techniques. We'll cover:

- What descriptive statistics are and why they matter
- Getting to know a dataset's structure
- Measures of central tendency (mean, median, mode)
- Measures of spread and variability (variance, standard deviation, range, IQR)
- Frequency tables and proportions for categorical data
- Using `summary()`, `skimr::skim()`, and `psych::describe()` for fast overviews
- Grouped descriptive statistics with `dplyr`
- Correlation between variables
- Visualizing distributions with `ggplot2`
- Handling missing data during descriptive analysis
- Putting it all together in a repeatable workflow

We'll use two datasets that ship with base R and ggplot2 — `mtcars` and `diamonds` — so you can follow along without downloading anything. Let's get started.

## What Is Descriptive Analysis, Really?

Descriptive analysis is the branch of statistics concerned with summarizing and organizing data so that it can be understood at a glance, without drawing conclusions that go beyond the data itself. It stands in contrast to **inferential statistics**, which uses sample data to make claims about a broader population — think confidence intervals, hypothesis tests, and regression coefficients with p-values attached.

Descriptive statistics don't try to predict anything or generalize to a population; they simply describe what's in front of you. If you calculate the average commute time for 200 survey respondents, that's descriptive. If you use that average to estimate the average commute time for an entire city, and attach a margin of error to it, that's inferential.

This distinction matters because descriptive analysis is a prerequisite for almost everything else you'll do with data. You cannot sensibly interpret a regression coefficient if you don't know the scale, spread, or distribution of the underlying variable. You cannot spot a data quality issue if you never look at minimums, maximums, and counts of missing values. Descriptive analysis is the seatbelt of data science — unglamorous, but the thing that keeps you from a serious accident.

Broadly, descriptive analysis covers four categories of tools:

1. **Measures of central tendency** — where is the "middle" of the data? (mean, median, mode)
2. **Measures of dispersion** — how spread out is the data? (range, variance, standard deviation, interquartile range)
3. **Measures of shape** — is the distribution symmetric, skewed, peaked, or flat? (skewness, kurtosis)
4. **Frequency and proportion summaries** — for categorical variables, how often does each category occur?

R gives you fast, idiomatic ways to compute all four categories, and we'll go through each one in turn.

## Getting to Know Your Dataset's Structure

Before computing a single statistic, you need to understand the shape and structure of your dataset: how many observations, how many variables, and what type each variable is (numeric, character, factor, logical, date). R provides several functions for this reconnaissance step.

```{r}
#| label: setup

library(dplyr)
library(ggplot2)

data(mtcars)
str(mtcars)
```

The `str()` function is often the very first thing an experienced R user runs on new data. It reports the number of observations and variables, and for each column, its data type and a preview of the first few values. In `mtcars`, every column happens to be numeric, but in most real-world datasets you'll see a mix of numeric, character, and factor columns.

Two close cousins of `str()` are `dim()`, which returns the number of rows and columns, and `glimpse()` from `dplyr`, which produces a similar (arguably more readable) summary:

```{r}
dim(mtcars)
glimpse(mtcars)
```

It's also worth checking column names and the first and last few rows of the data, using `names()`, `head()`, and `tail()`:

```{r}
names(mtcars)
head(mtcars, 3)
tail(mtcars, 3)
```

This might feel like a trivial step, but it catches an enormous number of problems early: columns that were imported as text instead of numbers, unexpected extra columns from a bad CSV parse, or a dataset that's smaller than you expected because a join dropped rows silently. Get in the habit of running `str()` or `glimpse()` immediately after loading any dataset, before you do anything else.

## Measures of Central Tendency

Central tendency statistics summarize where the "center" of a distribution sits. The three classic measures are the mean, median, and mode, and each tells you something slightly different.

### The Mean

The arithmetic mean is the sum of all values divided by the number of values. In R, it's computed with `mean()`:

```{r}
mean(mtcars$mpg)
```

The average fuel economy across the 32 cars in `mtcars` is about 20 miles per gallon. The mean is sensitive to every single value in the dataset, including outliers — a single car with extraordinarily high or low mpg will pull the mean in its direction. This sensitivity is both the mean's strength (it uses all available information) and its weakness (it can be misleading in skewed distributions).

### The Median

The median is the middle value when data is sorted — the point at which half the observations fall below and half fall above. Unlike the mean, the median is robust to outliers, which is why it's often preferred for skewed data like income or house prices.

```{r}
median(mtcars$mpg)
```

When the mean and median diverge noticeably, that's a signal the distribution is skewed. If the mean is higher than the median, the data has a long right tail (a few very large values are pulling the average up); if the mean is lower than the median, the tail is on the left.

### The Mode

Oddly, base R has no built-in `mode()` function for statistical mode (the `mode()` function in R actually refers to the storage mode of an object, a completely different concept). To find the most frequently occurring value, you typically write a small helper function or use a table:

```{r}
get_mode <- function(x) {
  freq_table <- table(x)
  names(freq_table)[which.max(freq_table)]
}

get_mode(mtcars$cyl)
```

The mode is most useful for categorical or discrete variables. In `mtcars`, the number of cylinders is a discrete variable, and the mode tells us the most common engine configuration among these cars. For continuous variables like mpg, the mode is less meaningful unless you're describing a specific peak in the distribution (in which case a density plot is usually more informative than a single mode value).

### Trimmed Means and Robust Alternatives

When your data contains outliers but you still want something mean-like, a **trimmed mean** — which discards a fixed percentage of the most extreme values before averaging — is a useful middle ground between the mean and median:

```{r}
mean(mtcars$mpg, trim = 0.1)
```

This computes the mean after removing the top and bottom 10% of values, reducing the influence of extreme observations while still using more of the distribution's information than the median alone.

## Measures of Dispersion

Central tendency tells you where the data sits, but it says nothing about how spread out the values are. Two datasets can have identical means and completely different variability — one tightly clustered, one wildly dispersed. Measures of dispersion capture this.

### Range

The range is simply the difference between the maximum and minimum values:

```{r}
range(mtcars$mpg)
diff(range(mtcars$mpg))
```

The `range()` function returns both the minimum and maximum; wrapping it in `diff()` gives you the single numeric range. The range is intuitive but fragile — it depends entirely on the two most extreme observations, which makes it very sensitive to outliers.

### Variance and Standard Deviation

Variance measures the average squared deviation from the mean, and standard deviation is simply the square root of variance, which brings the units back to the original scale of the data (making it much easier to interpret).

```{r}
var(mtcars$mpg)
sd(mtcars$mpg)
```

A standard deviation of roughly 6 mpg tells us that fuel economy values typically deviate from the mean of about 20 mpg by around 6 mpg in either direction. Standard deviation is one of the most widely used dispersion statistics because it plays nicely with the normal distribution and countless downstream techniques (z-scores, confidence intervals, control charts, and so on).

One subtlety worth knowing: R's `var()` and `sd()` compute the **sample** variance and standard deviation by default, dividing by `n - 1` rather than `n`. This is Bessel's correction, which produces an unbiased estimator of the population variance when working with a sample. It's the standard behavior for most statistical software, but worth remembering if you're comparing results against manual calculations.

### Interquartile Range (IQR)

The interquartile range measures the spread of the middle 50% of the data — the distance between the 75th percentile (Q3) and the 25th percentile (Q1). It's a robust measure of spread, largely unaffected by outliers.

```{r}
IQR(mtcars$mpg)
quantile(mtcars$mpg)
```

The `quantile()` function by default returns the minimum, Q1, median, Q3, and maximum — the five-number summary that underlies every boxplot you've ever seen. You can request arbitrary percentiles too:

```{r}
quantile(mtcars$mpg, probs = c(0.1, 0.9))
```

This tells you the 10th and 90th percentile values, useful when you want to describe the "typical range" of a variable while excluding the most extreme 10% on each end.

### Coefficient of Variation

When you want to compare the relative variability of two variables measured on different scales, the standard deviation alone isn't a fair comparison. The **coefficient of variation** (CV) — standard deviation divided by the mean, often expressed as a percentage — normalizes for scale:

```{r}
cv <- function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) * 100
cv(mtcars$mpg)
cv(mtcars$hp)
```

This tells you that horsepower varies more, relative to its own average, than mpg does — a comparison that raw standard deviations alone couldn't make fairly, since horsepower and mpg live on very different numeric scales.

## Skewness and Kurtosis

Beyond center and spread, it's often useful to describe the **shape** of a distribution. Two statistics capture this: skewness and kurtosis.

**Skewness** measures asymmetry. A skewness of zero indicates a symmetric distribution; positive skewness indicates a long right tail (common in income data, for example); negative skewness indicates a long left tail.

**Kurtosis** measures "tailedness" — how much of the distribution's variance comes from extreme, rare values versus frequent, moderate deviations. A normal distribution has a kurtosis of 3 (or 0 if using "excess kurtosis," which subtracts 3 as a baseline).

Base R doesn't include skewness or kurtosis functions, but the `moments` and `psych` packages do:

```{r}
library(psych)
skew(mtcars$mpg)
kurtosi(mtcars$mpg)
```

A positive skewness value here confirms that the mpg distribution has a longer tail toward higher values — consistent with a handful of very fuel-efficient cars pulling the mean upward relative to the median.

## Fast, Comprehensive Overviews

Computing each statistic one at a time is instructive when you're learning, but in practice you'll usually want a fast, comprehensive overview of an entire dataset. R offers several options, each with a different flavor.

### `summary()`

The base R workhorse. Calling `summary()` on a data frame produces the five-number summary (plus the mean) for every numeric column, and frequency counts for factors:

```{r}
summary(mtcars)
```

In a handful of lines, you get the minimum, first quartile, median, mean, third quartile, and maximum for every single numeric variable in the dataset. This is usually the very first command I run on any new dataset after `str()`.

### `psych::describe()`

The `psych` package's `describe()` function goes further, adding standard deviation, skewness, kurtosis, and standard error to the mix, and organizing everything into a tidy, easy-to-read table:

```{r}
describe(mtcars)
```

This single call replaces dozens of individual function calls and is often my preferred starting point when I want a genuinely thorough first look at a numeric dataset.

### `skimr::skim()`

The `skimr` package takes a more modern, visually enhanced approach. It works across mixed data types (numeric, character, factor, date) and even includes a miniature inline histogram for each numeric variable, letting you spot skew or bimodality without plotting a single chart:

```{r}
library(skimr)
skim(mtcars)
```

`skimr` is especially handy in exploratory workflows and R Markdown/Quarto reports because its output is designed to render cleanly, and it automatically groups summary statistics by variable type — a huge convenience once your dataset contains a mix of numeric, categorical, and logical columns.

## Descriptive Statistics for Categorical Data

Everything so far has focused on numeric variables, but categorical variables need their own toolkit: frequency tables, proportions, and cross-tabulations rather than means and standard deviations.

### Frequency Tables

The base `table()` function counts occurrences of each category:

```{r}
table(mtcars$cyl)
```

This tells you at a glance how many cars in the dataset have 4, 6, or 8 cylinders. To express these counts as proportions rather than raw counts, wrap the table in `prop.table()`:

```{r}
prop.table(table(mtcars$cyl))
```

Multiplying by 100 turns these into readable percentages:

```{r}
round(prop.table(table(mtcars$cyl)) * 100, 1)
```

### Cross-Tabulations

When you want to understand the relationship between two categorical variables, a two-way frequency table (cross-tabulation) is the natural tool:

```{r}
table(mtcars$cyl, mtcars$am)
```

This shows the joint distribution of cylinder count and transmission type (0 = automatic, 1 = manual) — how many 4-, 6-, and 8-cylinder cars have automatic versus manual transmissions. Adding margins gives you row and column totals:

```{r}
addmargins(table(mtcars$cyl, mtcars$am))
```

Cross-tabulations like this are often the first step toward more formal tests of association, such as the chi-squared test, but even on their own they're an excellent way to spot patterns and imbalances in categorical data.

## Grouped Descriptive Statistics with dplyr

Real analytical questions are rarely about a whole dataset in isolation — they're about how a statistic differs across groups. "What's the average fuel economy, broken down by number of cylinders?" is a far more useful question than "what's the average fuel economy?" alone. This is where `dplyr`'s `group_by()` and `summarize()` combination shines.

```{r}
mtcars |>
  group_by(cyl) |>
  summarize(
    n = n(),
    mean_mpg = mean(mpg),
    median_mpg = median(mpg),
    sd_mpg = sd(mpg),
    min_mpg = min(mpg),
    max_mpg = max(mpg)
  )
```

In a single, readable pipeline, we've computed the count, mean, median, standard deviation, minimum, and maximum fuel economy for each cylinder group. The pattern is always the same: group the data by one or more categorical variables with `group_by()`, then compute whatever summary statistics you need inside `summarize()`.

This generalizes naturally to multiple grouping variables:

```{r}
mtcars |>
  group_by(cyl, am) |>
  summarize(
    n = n(),
    mean_mpg = mean(mpg),
    .groups = "drop"
  )
```

Now we can see average fuel economy broken down simultaneously by cylinder count and transmission type — a much richer picture than either variable alone provides. The `.groups = "drop"` argument simply prevents dplyr from leaving the result grouped after the summary, which avoids surprises in any code that follows.

You can also combine `summarize()` with `across()` to compute the same statistic over many columns at once, which is invaluable when you have dozens of numeric variables and don't want to type out each one by hand:

```{r}
mtcars |>
  group_by(cyl) |>
  summarize(across(c(mpg, hp, wt), mean, .names = "mean_{col}"))
```

This single call computes the mean of `mpg`, `hp`, and `wt`, grouped by cylinder count, and gives each resulting column a clear, descriptive name.

## Correlation Analysis

Descriptive analysis isn't limited to single variables in isolation; understanding how variables relate to one another is just as important. The most common tool for this is the correlation coefficient, which measures the strength and direction of a linear relationship between two numeric variables.

```{r}
cor(mtcars$mpg, mtcars$wt)
```

A correlation of roughly -0.87 tells us that heavier cars tend to have substantially lower fuel economy — a strong negative linear relationship. Correlation coefficients range from -1 (perfect negative relationship) to +1 (perfect positive relationship), with 0 indicating no linear relationship at all.

For an entire dataset, `cor()` can compute a full correlation matrix in one call:

```{r}
round(cor(mtcars[, c("mpg", "hp", "wt", "qsec")]), 2)
```

This matrix shows every pairwise correlation among four variables at once. Scanning it quickly reveals that horsepower and weight are both strongly negatively correlated with fuel economy, while quarter-mile time (`qsec`) has a milder relationship with the others.

A quick caution: correlation only captures *linear* relationships, and it says nothing about causation. Two variables can be strongly correlated because one causes the other, because both are driven by a third factor, or by pure coincidence. Descriptive correlation analysis tells you *that* a relationship exists in your data — it never tells you *why*.

## Visualizing Distributions

Numbers alone can only take you so far — the human eye is remarkably good at spotting patterns, skew, outliers, and multimodal distributions that summary statistics can hide. Pairing every descriptive statistic with a visualization is a habit worth building early. `ggplot2` is the natural companion here.

### Histograms

Histograms show the distribution of a single numeric variable by bucketing values into bins and counting observations in each:

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Fuel Economy", x = "Miles per Gallon", y = "Count") +
  theme_minimal()
```

Histograms reveal skewness, gaps, and multimodality that a single mean and standard deviation would never show you. A dataset with two distinct peaks, for instance, might indicate two underlying subgroups that should be analyzed separately rather than pooled together.

### Boxplots

Boxplots are a compact visual encoding of the five-number summary (minimum, Q1, median, Q3, maximum), plus any points flagged as outliers:

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Fuel Economy by Cylinder Count", x = "Cylinders", y = "Miles per Gallon") +
  theme_minimal()
```

Boxplots are especially powerful when comparing a numeric variable across several categorical groups, since you can quickly compare medians, spread, and the presence of outliers side by side.

### Bar Charts for Categorical Data

For categorical variables, a bar chart of counts or proportions is the visual equivalent of a frequency table:

```{r}
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "coral") +
  labs(title = "Count of Cars by Cylinder Count", x = "Cylinders", y = "Count") +
  theme_minimal()
```

### Scatter Plots for Relationships

To visualize the relationship behind a correlation coefficient, a scatter plot is indispensable:

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "darkgreen", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Fuel Economy vs. Weight", x = "Weight (1000 lbs)", y = "Miles per Gallon") +
  theme_minimal()
```

Adding a fitted trend line with `geom_smooth()` gives an immediate visual sense of the linear relationship that the correlation coefficient described numerically, and it also makes any nonlinearity or heteroscedasticity in the relationship easy to spot — something a single correlation number cannot reveal.

## Handling Missing Data in Descriptive Analysis

Real-world data almost always contains missing values, and they complicate descriptive analysis in subtle ways. Most base R statistical functions return `NA` if any value in the input is missing, which is R's way of forcing you to make a deliberate decision about how to handle them rather than silently ignoring the problem.

```{r}
mtcars_na <- mtcars
mtcars_na$mpg[c(2, 5)] <- NA

mean(mtcars_na$mpg)
mean(mtcars_na$mpg, na.rm = TRUE)
```

The first call returns `NA` because two values are missing; the second, with `na.rm = TRUE`, excludes missing values from the calculation and returns a usable result. Nearly every descriptive statistics function in R — `mean()`, `median()`, `sd()`, `var()`, `sum()`, `min()`, `max()` — accepts an `na.rm` argument for exactly this reason.

Before simply dropping missing values, though, it's worth quantifying how much missingness you're dealing with, and whether it's concentrated in particular rows, columns, or subgroups:

```{r}
colSums(is.na(mtcars_na))
mean(is.na(mtcars_na$mpg)) * 100
```

The first line counts missing values per column; the second expresses the proportion of missing values in a single column as a percentage. If missingness is substantial or systematic — for example, if a survey question was skipped disproportionately by one demographic group — that pattern itself is an important descriptive finding, often more important than the summary statistic you were originally trying to compute. Tools like `skimr::skim()` report missingness alongside every other summary statistic automatically, which is one more reason it's a great default choice for a first pass over new data.

## A Repeatable Descriptive Analysis Workflow

Pulling everything together, here's a workflow I'd recommend for any new dataset that lands in your lap:

1. **Inspect structure.** Run `str()`, `dim()`, and `glimpse()` to understand size, shape, and variable types.
2. **Check for missingness.** Use `colSums(is.na(df))` or `skimr::skim(df)` to see how much data is missing and where.
3. **Summarize numeric variables.** Use `summary()`, `psych::describe()`, or `skimr::skim()` for a fast overview of central tendency, spread, and shape.
4. **Summarize categorical variables.** Use `table()` and `prop.table()` for frequencies and proportions.
5. **Break down by group.** Use `dplyr::group_by()` and `summarize()` to see how key statistics vary across meaningful subgroups.
6. **Check relationships.** Use `cor()` for numeric variable pairs and cross-tabulations for categorical pairs.
7. **Visualize.** Pair every important statistic with a histogram, boxplot, bar chart, or scatter plot — numbers hide things that pictures reveal.

Following these seven steps in order, every time, before you touch a model or a hypothesis test, will save you from an enormous number of downstream headaches: it catches data entry errors, reveals skewed variables that might need transformation, flags outliers that could distort a model, and builds the kind of intuition about your data that no algorithm can substitute for.

## Putting It All Together: A Worked Example

Let's apply this entire workflow, start to finish, to a slightly richer dataset — the `diamonds` dataset from `ggplot2`, which contains prices and attributes for about 54,000 diamonds.

```{r}
data(diamonds)
glimpse(diamonds)
```

Structure check complete — we have roughly 54,000 rows and 10 columns, a mix of numeric measurements (carat, depth, table, price, and physical dimensions) and ordered factors (cut, color, clarity).

```{r}
colSums(is.na(diamonds))
```

No missing values, which is good news and lets us move straight to summarizing.

```{r}
summary(diamonds$price)
sd(diamonds$price)
```

The mean price is noticeably higher than the median, and the standard deviation is large relative to the mean — a strong hint that price is right-skewed, likely due to a small number of very expensive diamonds. Let's confirm visually:

```{r}
ggplot(diamonds, aes(x = price)) +
  geom_histogram(bins = 50, fill = "purple", alpha = 0.7) +
  labs(title = "Distribution of Diamond Prices", x = "Price (USD)", y = "Count") +
  theme_minimal()
```

The histogram confirms the skew we suspected from the summary statistics alone — a long tail of expensive diamonds pulling the mean above the median. This kind of skew is extremely common in price, income, and other economic data, and it's exactly the sort of thing that would be missed if we only looked at the mean and never plotted the distribution.

Next, let's see how price varies by diamond cut quality:

```{r}
diamonds |>
  group_by(cut) |>
  summarize(
    n = n(),
    mean_price = mean(price),
    median_price = median(price),
    .groups = "drop"
  )
```

Interestingly, "Fair" cut diamonds have a higher average price than "Ideal" cut diamonds — a counterintuitive result at first glance, until you remember that carat weight (a major price driver) isn't held constant here. This is a perfect illustration of why descriptive analysis is a starting point for inquiry, not an endpoint: the finding raises a question ("why would lower-quality cuts fetch higher prices?") that deserves further investigation, likely by examining carat weight within each cut category.

```{r}
diamonds |>
  group_by(cut) |>
  summarize(mean_carat = mean(carat), .groups = "drop")
```

Sure enough, "Fair" diamonds tend to be larger on average, which explains the apparent paradox — larger stones command higher prices regardless of cut quality. This single follow-up query, uncovered through nothing more than grouped descriptive statistics, would have been completely invisible if we'd stopped at the overall summary of price.

Finally, let's check the correlation between carat and price:

```{r}
cor(diamonds$carat, diamonds$price)
```

A correlation above 0.9 confirms that carat weight is an extremely strong predictor of price — unsurprising, but useful to confirm numerically before building any predictive model. A scatter plot makes the (strongly nonlinear) relationship even clearer:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, color = "darkblue") +
  labs(title = "Diamond Price vs. Carat Weight", x = "Carat", y = "Price (USD)") +
  theme_minimal()
```

Notice the curved, fan-shaped relationship rather than a straight line — a detail the correlation coefficient alone (which only measures linear association) doesn't fully capture. This is exactly why pairing every numeric summary with a visualization matters: the correlation confirms a strong relationship exists, but only the scatter plot reveals its true shape.

## Common Pitfalls to Avoid

A few mistakes come up again and again in descriptive analysis, and it's worth calling them out explicitly:

**Relying on the mean alone.** The mean is only a good summary of central tendency when the data is roughly symmetric. For skewed distributions (income, price, wait times), the median is usually more representative, and reporting both is safer than reporting either alone.

**Ignoring missing data.** Silently dropping rows with `na.rm = TRUE` without first checking how much data is missing, or whether the missingness follows a pattern, can quietly bias every subsequent statistic.

**Skipping visualization.** Two datasets can share an identical mean, median, and standard deviation while looking completely different when plotted — this is the entire point of Anscombe's quartet, a classic statistical example designed to demonstrate exactly this. Never trust summary statistics without a corresponding plot.

**Confusing correlation with causation.** A strong correlation between two variables never, on its own, establishes that one causes the other.

**Treating factors as numbers (or vice versa).** Computing a mean on a numerically-coded categorical variable (like a survey response coded 1–5) can be misleading if the categories aren't genuinely ordered on an equal-interval scale.

**Not checking group sizes.** A grouped summary statistic computed from a group of 3 observations carries far less weight than one computed from a group of 3,000 — always report (or at least check) the `n` behind every grouped mean or proportion.

## Frequently Asked Questions

**Do I need to learn base R functions if I already know the tidyverse?**

Yes, at least the essentials. Functions like `mean()`, `median()`, `sd()`, `summary()`, and `table()` are used everywhere, including inside `dplyr::summarize()` calls, and they show up constantly in other people's code, package documentation, and Stack Overflow answers. The tidyverse and base R aren't competitors here — `group_by()` and `summarize()` are simply a more expressive way of applying the same base functions across groups.

**Should I always report both mean and median?**

For any variable where skewness is plausible — money, time, counts, ratings — yes. It costs one extra line of code and immediately tells your reader (or future self) whether the distribution is symmetric or skewed, without requiring them to trust a single number in isolation.

**Is `skimr` better than `summary()` and `psych::describe()`?**

They're complementary rather than competing. `summary()` is built in and requires no extra packages, which makes it a safe default in any environment. `psych::describe()` adds skewness, kurtosis, and standard error in a clean table. `skimr::skim()` adds missing-value counts and inline mini-histograms and handles mixed data types most gracefully. In practice, many analysts reach for `skimr::skim()` first for a broad overview, then drop into `summary()` or `describe()` for a closer look at specific variables.

**How much descriptive analysis is "enough" before moving on to modeling?**

There's no fixed rule, but a reasonable minimum is: know the size and structure of your dataset, know which variables have missing data and how much, have a summary statistic and a plot for every variable you plan to use in a model, and understand the basic pairwise relationships between your key variables. If you can't answer "what does this variable typically look like, and how does it relate to my outcome of interest?" for every variable in your model, it's worth spending more time here before proceeding.

**Can descriptive statistics be misleading?**

Absolutely — that's precisely why this article emphasizes pairing every statistic with a corresponding visualization, and why measures like the mean, correlation coefficient, and standard deviation each come with important caveats about what they can and cannot tell you. Descriptive statistics summarize; they don't explain, and it's easy to over-interpret a clean-looking number that's hiding a messy underlying distribution.

## Conclusion

Descriptive analysis rarely gets the spotlight in data science curricula or conference talks — that attention usually goes to machine learning, deep learning, or the newest inferential technique. But every experienced analyst will tell you the same thing: the quality of everything downstream depends entirely on how well you understood your data in the first place. A model built on a dataset you never properly explored is a model built on sand.

R makes this exploratory step genuinely enjoyable. A handful of functions — `str()`, `summary()`, `table()`, `group_by()`/`summarize()`, `cor()` — combined with `ggplot2` for visualization, give you everything you need to build real intuition about any dataset, in minutes rather than hours. Packages like `skimr` and `psych` go even further, packaging dozens of useful statistics into a single, readable call.

The next time a new dataset lands in your inbox, resist the urge to jump straight to modeling. Spend fifteen minutes with `str()`, `summary()`, a few `group_by()` calls, and a handful of histograms and boxplots. You'll be amazed how often that fifteen minutes saves you hours of confusion — or an embarrassing mistake — later on.

---

*Have questions about descriptive analysis in R, or want to see a follow-up article on inferential statistics or hypothesis testing? Let me know in the comments.*
