---
title: "Making Forest Plots in R: A Practical Guide"
description: "A hands-on walkthrough of building forest plots in R using ggplot2, the forestplot package, and metafor — from simple effect-size charts to publication-ready meta-analysis figures."
author: "Your Name"
date: today
categories: [R, data-visualization, statistics, meta-analysis]
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
---

## Introduction

If you have ever opened a meta-analysis paper, a clinical trial report, or a subgroup analysis in an epidemiology study, you have almost certainly seen a forest plot. It is one of those visualizations that looks simple at first glance — a column of dots with horizontal lines running through them — but that simplicity is exactly what makes it so effective. A forest plot lets a reader scan dozens of studies, subgroups, or model coefficients in seconds and immediately see which ones point in which direction, how precise each estimate is, and where the overall summary effect lands.

Despite how common forest plots are in the applied statistics and epidemiology literature, a lot of people find them oddly fiddly to build the first time they try. Part of the reason is that there are several genuinely different ways to make one in R, and each comes from a different corner of the ecosystem with its own conventions: `forestplot` is built for exactly this chart type and nothing else, `ggplot2` is a general-purpose grammar-of-graphics package that can be coaxed into producing a forest plot with `geom_pointrange()`, and `metafor` is a full meta-analysis package that happens to include a forest-plot function tailored to its own model objects.

This post walks through all three approaches. We will start with the conceptual anatomy of a forest plot, move into a from-scratch `ggplot2` version so you understand exactly what each visual element represents, then look at the dedicated `forestplot` package for when you want fine control over a table-plus-chart layout, and finish with `metafor`, which is the natural choice if you are actually running a meta-analysis rather than just visualizing one. Along the way we will cover common customizations: log scales, custom labels, grouping and stratification, confidence interval styling, and adding summary diamonds.

By the end, you should be able to pick the right tool for your situation and build a forest plot that looks like it belongs in a journal, not like a default `ggplot()` call.

## What Exactly Is a Forest Plot?

A forest plot is fundamentally a visual table. Each row represents one "unit" — a study, a subgroup, a regression coefficient, a treatment arm comparison — and shows three pieces of information for that unit:

1. **A point estimate**, usually a mean difference, odds ratio, risk ratio, hazard ratio, or standardized effect size, plotted as a dot, square, or diamond.
2. **A confidence interval**, plotted as a horizontal line (or "whisker") extending from the point estimate, showing the precision of that estimate.
3. **A reference line**, usually at zero (for differences) or one (for ratios), representing "no effect," so readers can see at a glance whether each interval crosses the null.

Frequently, forest plots also include:

- A **summary diamond** at the bottom representing the pooled or overall estimate, often from a fixed-effect or random-effects meta-analysis.
- **Weights**, shown either as the size of the point marker or as a column of numbers, reflecting how much each study contributes to the pooled estimate (larger or more precise studies get more weight).
- A **table of numbers** to the left or right of the plot showing sample sizes, event counts, or the numeric estimate and CI as text, since eyeballing a dot on a log scale is not always precise enough.

The reason forest plots work so well is that they let you compare *both* the direction and the precision of many estimates simultaneously, something a table of numbers alone cannot do nearly as efficiently. A study with a wide confidence interval visually announces its own uncertainty, while a study with a tight interval draws the eye as a more reliable data point — even before a reader consciously registers the sample size column.

## Setting Up

We will use a handful of packages throughout this post. If you do not already have them installed, run:

```{r}
#| eval: false
install.packages(c("ggplot2", "dplyr", "forestplot", "metafor", "scales"))
```

And load what we need:

```{r}
library(ggplot2)
library(dplyr)
library(scales)
```

We will load `forestplot` and `metafor` in their respective sections, since both packages export functions with generic names (like `forestplot()`) that can conflict with each other, and it is good practice to be explicit about where a function comes from when that happens.

## A Toy Dataset

To keep the examples concrete, let's imagine a meta-analysis of eight hypothetical clinical trials testing a new drug against a placebo for reducing hospital readmission risk. Each trial reports an odds ratio, along with a 95% confidence interval, and a study weight based on inverse-variance.

```{r}
studies <- data.frame(
  study = c("Anderson 2015", "Baxter 2016", "Chen 2017", "Diaz 2018",
            "Evans 2019", "Fischer 2020", "Garcia 2021", "Hughes 2022"),
  n = c(120, 340, 88, 512, 76, 205, 640, 150),
  or = c(0.72, 0.85, 1.15, 0.68, 1.40, 0.91, 0.77, 0.60),
  lower = c(0.48, 0.63, 0.72, 0.51, 0.85, 0.66, 0.62, 0.35),
  upper = c(1.08, 1.15, 1.83, 0.90, 2.30, 1.26, 0.96, 1.03),
  weight = c(9.5, 15.2, 6.1, 18.4, 4.8, 11.3, 22.0, 7.7)
)

studies
```

Each row is a study. The `or` column is the point estimate (an odds ratio, so 1.0 means "no effect"), and `lower`/`upper` are the bounds of the 95% confidence interval. The `weight` column reflects how much influence each study should have in a pooled estimate — larger, more precise studies get bigger weights.

We also want a summary row representing the pooled estimate across all eight studies. For now we will just compute a simple inverse-variance-weighted average on the log scale (this is roughly what a fixed-effect meta-analysis does under the hood); later, in the `metafor` section, we will let the package do this properly.

```{r}
log_or <- log(studies$or)
se <- (log(studies$upper) - log(studies$lower)) / (2 * 1.96)
w <- 1 / se^2

pooled_log_or <- sum(log_or * w) / sum(w)
pooled_se <- sqrt(1 / sum(w))

pooled_or <- exp(pooled_log_or)
pooled_lower <- exp(pooled_log_or - 1.96 * pooled_se)
pooled_upper <- exp(pooled_log_or + 1.96 * pooled_se)

c(pooled_or = pooled_or, pooled_lower = pooled_lower, pooled_upper = pooled_upper)
```

Now we have everything we need for our first plot.

## Building a Forest Plot from Scratch with ggplot2

There is real value in building one of these by hand at least once, because it forces you to understand exactly what a forest plot is made of: it is nothing more than a `geom_pointrange()` (or the combination of `geom_point()` and `geom_errorbarh()`) plotted against a categorical y-axis, with a vertical reference line and some axis transformations.

### The Basic Version

```{r}
ggplot(studies, aes(x = or, y = reorder(study, or))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  labs(
    x = "Odds Ratio (95% CI)",
    y = NULL,
    title = "Effect of Drug X on Hospital Readmission"
  ) +
  theme_minimal()
```

That is already a recognizable forest plot: one row per study, a dot for the point estimate, a horizontal line for the interval, and a dashed reference line at the null value of 1. A few things to notice about the code:

- `geom_pointrange()` needs `xmin` and `xmax` aesthetics in addition to `x` and `y` — it draws the point at `x` and the range from `xmin` to `xmax`.
- We used `reorder(study, or)` to sort studies by their effect size rather than relying on alphabetical or input order. Depending on your audience, you might instead want to preserve chronological order (by publication year) or group order (by subgroup), so treat this as a default rather than a rule.
- Because odds ratios are inherently multiplicative rather than additive, this axis should almost always be shown on a **log scale**, which we have not done yet — that is next.

### Putting the Effect-Size Axis on a Log Scale

For ratio measures (odds ratios, risk ratios, hazard ratios), a linear x-axis is misleading: the visual distance between an OR of 0.5 and 1.0 should be the same as between 1.0 and 2.0, since both represent a "doubling" in one direction or the other. `scale_x_log10()` handles this for us.

```{r}
ggplot(studies, aes(x = or, y = reorder(study, or))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4)) +
  labs(
    x = "Odds Ratio (95% CI, log scale)",
    y = NULL,
    title = "Effect of Drug X on Hospital Readmission"
  ) +
  theme_minimal()
```

Notice how the intervals now look more symmetric around their point estimates — that symmetry on the log scale is actually a sign the plot is behaving correctly, since odds ratios are log-normally distributed in most standard models.

### Sizing Points by Study Weight

A common refinement is to scale the size of each point by the study's weight in the pooled analysis, so heavier studies visually dominate the plot the same way they dominate the underlying statistics.

```{r}
ggplot(studies, aes(x = or, y = reorder(study, or))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.15, color = "grey30") +
  geom_point(aes(size = weight), shape = 15, color = "steelblue4") +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4)) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  labs(
    x = "Odds Ratio (95% CI, log scale)",
    y = NULL,
    title = "Effect of Drug X on Hospital Readmission",
    subtitle = "Point size reflects study weight"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Here we switched from `geom_pointrange()` to the combination of `geom_errorbarh()` and `geom_point()`, since `geom_pointrange()` does not give you independent control over point size versus line appearance. We also switched to square markers (`shape = 15`), which is a common convention for individual study estimates, reserving diamonds for pooled summary estimates.

### Adding the Summary Diamond

To show a pooled estimate, we typically draw a diamond whose horizontal extent represents the confidence interval and whose vertical points represent... well, nothing statistically meaningful, but it is the conventional shape that signals "this is a summary, not an individual study." We can build a diamond manually with `geom_polygon()`.

```{r}
summary_row <- data.frame(study = "Pooled Estimate")

diamond <- data.frame(
  x = c(pooled_lower, pooled_or, pooled_upper, pooled_or),
  y = c(0, 0.3, 0, -0.3)
)

plot_data <- studies %>%
  mutate(study = factor(study, levels = c(rev(study[order(studies$or)]), "Pooled Estimate")))

ggplot() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_errorbarh(
    data = studies,
    aes(xmin = lower, xmax = upper,
        y = factor(study, levels = c("Pooled Estimate", rev(study[order(studies$or)])))),
    height = 0.15, color = "grey30"
  ) +
  geom_point(
    data = studies,
    aes(x = or, y = factor(study, levels = c("Pooled Estimate", rev(study[order(studies$or)]))),
        size = weight),
    shape = 15, color = "steelblue4"
  ) +
  geom_polygon(
    data = transform(diamond, y = y + 1),
    aes(x = x, y = y),
    fill = "firebrick4"
  ) +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4)) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  labs(
    x = "Odds Ratio (95% CI, log scale)",
    y = NULL,
    title = "Effect of Drug X on Hospital Readmission",
    subtitle = "Diamond shows the pooled fixed-effect estimate"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

This works, but as you can see, hand-positioning a diamond via `geom_polygon()` on a categorical axis is finicky — you are essentially faking a numeric y-position for a factor level. This is exactly the kind of bookkeeping that a dedicated forest plot package handles for you, which brings us to the next section.

## The `forestplot` Package

The `forestplot` package (by Max Gordon) was built specifically for this chart type, and its signature feature is that it renders the chart alongside a table of text columns — study names, sample sizes, numeric estimates — all row-aligned with the plot. This is the layout you see constantly in Cochrane reviews and clinical trial papers.

```{r}
library(forestplot)
```

### Basic Usage

The `forestplot` package expects your labels as a matrix or data frame of text (which becomes the left-hand table) and your numeric estimates as separate vectors for the mean, lower, and upper bounds.

```{r}
label_text <- cbind(
  c("Study", studies$study, "Pooled Estimate"),
  c("N", studies$n, sum(studies$n)),
  c("OR (95% CI)",
    sprintf("%.2f (%.2f–%.2f)", studies$or, studies$lower, studies$upper),
    sprintf("%.2f (%.2f–%.2f)", pooled_or, pooled_lower, pooled_upper))
)

mean_vec  <- c(NA, studies$or, pooled_or)
lower_vec <- c(NA, studies$lower, pooled_lower)
upper_vec <- c(NA, studies$upper, pooled_upper)

forestplot(
  labeltext = label_text,
  mean = mean_vec,
  lower = lower_vec,
  upper = upper_vec,
  is.summary = c(TRUE, rep(FALSE, nrow(studies)), TRUE),
  zero = 1,
  xlog = TRUE,
  boxsize = 0.15,
  col = fpColors(box = "steelblue4", line = "grey30", summary = "firebrick4"),
  xlab = "Odds Ratio"
)
```

A few things worth unpacking here:

- `labeltext` takes a **matrix** where each column becomes a text column in the table, and the first row is the header. We built it with `cbind()` so that "Study", "N", and "OR (95% CI)" line up with their respective columns.
- `mean`, `lower`, and `upper` must have the same length as the number of rows in `labeltext`, including the header row (hence the leading `NA`) and the summary row.
- `is.summary` is a logical vector marking which rows should be styled as headers/summaries (typically bold, and rendered as diamonds instead of boxes).
- `zero = 1` sets the null-effect reference line, and `xlog = TRUE` puts the effect axis on a log scale — both essential when your effect measure is a ratio.
- `fpColors()` lets you control the color of the individual-study boxes, the confidence interval lines, and the summary diamond independently.

### Why Use `forestplot` Instead of `ggplot2`?

The advantage of `forestplot` is that the row-aligned text table is a first-class feature, not something you have to hack together with `patchwork` or `cowplot` to glue a `ggplot2` table and chart side by side. If your journal or report requires the classic "study | n | events | weight | estimate (CI) | plot" layout, `forestplot` gets you there with far less code than assembling it manually in `ggplot2`.

The tradeoff is that `forestplot` uses base graphics under the hood (via the `grid` package), so if the rest of your report or dashboard is built in `ggplot2`/`patchwork`, mixing the two visual systems can feel inconsistent, and theming options are more limited than what `ggplot2` offers.

### Customizing Box Size by Weight

Just like we sized points by weight in `ggplot2`, `forestplot` accepts an `info` argument (sometimes passed via `boxsize`) that scales each box.

```{r}
forestplot(
  labeltext = label_text,
  mean = mean_vec,
  lower = lower_vec,
  upper = upper_vec,
  is.summary = c(TRUE, rep(FALSE, nrow(studies)), TRUE),
  zero = 1,
  xlog = TRUE,
  boxsize = c(NA, studies$weight / max(studies$weight) * 0.3, NA),
  col = fpColors(box = "darkorange3", line = "grey30", summary = "firebrick4"),
  xlab = "Odds Ratio",
  txt_gp = fpTxtGp(label = gpar(cex = 0.9), ticks = gpar(cex = 0.8))
)
```

Passing a vector to `boxsize` instead of a single number lets each row's box scale independently — here proportional to `weight`, giving the same visual weighting effect we built manually in `ggplot2`, but without the diamond polygon gymnastics.

## Forest Plots the "Real" Way: Using `metafor`

If you are actually conducting a meta-analysis — rather than visualizing effect sizes you already computed elsewhere — you should generally be using the `metafor` package to fit the model in the first place, and letting its built-in `forest()` function draw the plot. This has a major advantage: the plot is generated directly from the fitted model object, so there is no risk of a mismatch between the numbers you analyzed and the numbers you plotted.

```{r}
library(metafor)
```

### Fitting a Random-Effects Model

`metafor` works with effect sizes and their sampling variances. For our odds-ratio data, we will first convert to the log scale (which is what meta-analytic models should be fit on) using `escalc()`, and then fit a random-effects model with `rma()`.

```{r}
dat <- escalc(
  measure = "OR",
  ai = round(studies$n * 0.15),   # illustrative "event" counts, treatment arm
  n1i = studies$n,
  ci = round(studies$n * 0.20),   # illustrative "event" counts, control arm
  n2i = studies$n,
  data = studies,
  slab = study
)

# For this illustration we'll instead build directly from the OR/CI we already have,
# since we didn't start with raw 2x2 counts.
dat$yi <- log(studies$or)
dat$vi <- ((log(studies$upper) - log(studies$lower)) / (2 * 1.96))^2

res <- rma(yi, vi, data = dat, method = "REML", slab = studies$study)
summary(res)
```

Note the workaround here: in a real analysis you would typically start from raw counts (events and non-events in each arm) and let `escalc()` compute the log odds ratios and variances for you. Since our toy dataset started from already-summarized odds ratios and confidence intervals, we backed out the log-scale effect size (`yi`) and its variance (`vi`) directly — a common and perfectly valid situation when you are extracting already-published summary statistics for a meta-analysis rather than raw counts.

### The Default `forest()` Plot

With a fitted `rma` object in hand, producing a forest plot is a single line:

```{r}
forest(res, atransf = exp, xlab = "Odds Ratio", header = "Study")
```

A few things are happening automatically here that we had to build by hand earlier:

- `atransf = exp` tells `forest()` that the model was fit on the log scale, so it should back-transform the axis labels to the odds-ratio scale for display, while still plotting positions on the (additive, symmetric) log scale internally. This is the statistically correct way to show ratio measures — better than merely calling `scale_x_log10()` on already-exponentiated numbers, because the *underlying model's* estimation happened on the log scale.
- The summary diamond at the bottom is drawn automatically from the fitted `res$b` (pooled estimate) and its confidence interval.
- Study labels (`slab`) are pulled from the model object, so there's no separate label vector to keep in sync.
- Heterogeneity statistics (I², tau², Q-test) can be added directly to the plot with `addfit`, `header`, or by using `metafor`'s companion `robu`/`predict` utilities — something neither `ggplot2` nor `forestplot` gives you out of the box, because those packages don't know anything about the statistical model behind the numbers.

### Adding Heterogeneity Statistics

It's standard practice in published meta-analyses to annotate the plot with I² (the percentage of variability due to heterogeneity rather than chance) and the results of Cochran's Q test. `metafor` makes this straightforward:

```{r}
forest(
  res,
  atransf = exp,
  xlab = "Odds Ratio",
  header = "Study",
  mlab = bquote(paste(
    "RE Model (Q = ", .(formatC(res$QE, digits = 2, format = "f")),
    ", df = ", .(res$k - 1),
    ", p = ", .(formatC(res$QEp, digits = 3, format = "f")),
    "; ", I^2, " = ", .(formatC(res$I2, digits = 1, format = "f")), "%)"
  ))
)
```

The `mlab` argument replaces the default "RE Model" label under the summary diamond with a richer annotation built using `bquote()` for the mathematical formatting of I².

### Subgroup Forest Plots

A very common real-world need is to stratify studies by a categorical variable — say, by whether the trial was conducted in an inpatient or outpatient setting — and show subgroup-specific pooled estimates alongside the overall one. Suppose we add a `setting` column to our data:

```{r}
studies$setting <- c("Inpatient", "Outpatient", "Inpatient", "Outpatient",
                      "Inpatient", "Outpatient", "Inpatient", "Outpatient")

dat$setting <- studies$setting
```

We can fit separate models per subgroup and stitch them into one plot using `rows` and repeated calls to `forest()`/`addpoly()`, but the cleaner modern approach is to fit one `rma()` per subgroup and then combine with `metafor::forest()`'s `rows` argument:

```{r}
res_in  <- rma(yi, vi, data = dat, subset = (setting == "Inpatient"))
res_out <- rma(yi, vi, data = dat, subset = (setting == "Outpatient"))

forest(
  res,
  atransf = exp,
  xlab = "Odds Ratio",
  header = "Study",
  order = studies$setting,
  rows = c(2:5, 9:12) # example row placement leaving gaps for subgroup headers
)

# Add subgroup summary diamonds
addpoly(res_in,  row = 6.5, mlab = "Subgroup: Inpatient",  atransf = exp)
addpoly(res_out, row = 0.5, mlab = "Subgroup: Outpatient", atransf = exp)
```

This kind of layered construction — base plot, then `addpoly()` calls for subgroup and overall summaries — is exactly the workflow used in most published Cochrane-style forest plots, and it is very hard to reproduce in `ggplot2` without a lot of manual row-position bookkeeping (though it is doable, as we saw in the diamond example earlier). If subgroup meta-analysis is genuinely your task, this is the strongest argument for working in `metafor` from the start rather than exporting numbers into `ggplot2`.

## Choosing the Right Tool

At this point it's worth stepping back and summarizing when each approach makes sense, since the three tools we've covered solve genuinely different problems.

**Use `ggplot2`** when:

- You want full visual control and your forest plot needs to match a broader `ggplot2`-themed report or dashboard.
- The "study" axis isn't really studies at all — maybe it's regression coefficients, subgroup effects from a single model, or scenario comparisons — and you don't need any meta-analytic machinery (pooling, heterogeneity tests) behind it.
- You're comfortable doing a bit more manual layout work in exchange for consistency with the rest of your visualization pipeline.

**Use `forestplot`** when:

- You need the classic row-aligned table-plus-chart layout (study name, N, events, weight, numeric CI, and the plot itself, all row-matched) without stitching together `ggplot2` and a table package yourself.
- You already have summary numbers (from a paper, a spreadsheet, or another program like RevMan or Stata) and just need to visualize them — you are not fitting a model in R.
- Base-graphics styling is acceptable for your output format.

**Use `metafor`** when:

- You are actually running the meta-analysis in R — fitting fixed-effect or random-effects models, computing heterogeneity statistics, doing subgroup or meta-regression analyses — and want the plot to be generated directly and correctly from that model.
- You need annotations like I², tau², or Q-test results integrated into the plot.
- You want to guarantee that the numbers on the plot and the numbers in your statistical results always match, because they come from the same object.

A reasonable default I'd suggest: if your task is fundamentally "fit and report a meta-analysis," start in `metafor` and let its `forest()` function do the visualization — you get correctness and heterogeneity reporting for free. If your task is fundamentally "visualize some effect sizes I already have, and it needs to fit a specific report style or match other charts," reach for `forestplot` (for the classic clinical layout) or `ggplot2` (for maximal flexibility and consistency with other plots in the same document).

## Polishing: Small Details That Make a Big Difference

Once you've settled on a base approach, a handful of small refinements tend to separate a "good enough" forest plot from one that looks genuinely publication-ready.

### 1. Sort Meaningfully, Not Alphabetically

Sorting by effect size (as we did earlier with `reorder()`) makes trends easy to spot, but it isn't always appropriate — for a meta-analysis, convention is usually to sort chronologically by publication year, or to group by subgroup and then by year within subgroup, so readers can track how evidence accumulated over time. Choose the ordering that matches the story you're telling, not just the one that looks tidiest.

### 2. Be Deliberate About the x-axis Range

Default axis limits in `ggplot2` and `forestplot` will often be dictated by whichever confidence interval happens to be widest — which can compress all the other, more precise studies into a narrow visual band in the middle of the plot. It's often worth manually setting `xlim` (in `ggplot2`, via `coord_cartesian(xlim = ...)` so you don't accidentally clip data; in `forestplot`, via the `clip` argument) to a sensible round-number range like 0.25–4 for odds ratios, and noting in a caption if any interval extends beyond the visible range.

```{r}
ggplot(studies, aes(x = or, y = reorder(study, or))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4)) +
  coord_cartesian(xlim = c(0.25, 4)) +
  labs(x = "Odds Ratio (95% CI, log scale)", y = NULL) +
  theme_minimal()
```

### 3. Use Color Sparingly and Purposefully

It's tempting to color each study a different hue, but that adds visual noise without adding information, since color in a forest plot should encode something meaningful — subgroup membership, statistical significance, or study quality — rather than just differentiate rows that are already differentiated by their vertical position and label.

```{r}
studies <- studies %>%
  mutate(significant = ifelse(lower > 1 | upper < 1, "Significant", "Not significant"))

ggplot(studies, aes(x = or, y = reorder(study, or), color = significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4)) +
  scale_color_manual(values = c("Significant" = "firebrick4", "Not significant" = "grey40")) +
  labs(
    x = "Odds Ratio (95% CI, log scale)", y = NULL, color = NULL,
    title = "Effect of Drug X on Hospital Readmission"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### 4. Label the Reference Line

A dashed vertical line at 1 (or 0) is only self-explanatory to readers who already know how to read a forest plot. Adding a small annotation — "Favors Treatment" on one side, "Favors Control" on the other — makes the chart accessible to a much wider audience, and is standard practice in clinical journals.

```{r}
ggplot(studies, aes(x = or, y = reorder(study, or))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4)) +
  coord_cartesian(xlim = c(0.2, 5), clip = "off") +
  annotate("text", x = 0.3, y = 0.3, label = "Favors Treatment", hjust = 0, size = 3.2, color = "grey30") +
  annotate("text", x = 3.5, y = 0.3, label = "Favors Control", hjust = 1, size = 3.2, color = "grey30") +
  labs(x = "Odds Ratio (95% CI, log scale)", y = NULL) +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 25, 10))
```

### 5. Keep Text Legible When Exporting

Forest plots are often exported at small sizes for two-column journal layouts, and default `ggplot2` text sizes that look fine on screen can become unreadable at print resolution. When saving, it's worth explicitly setting `base_size` in your theme and testing the actual exported file at its true dimensions rather than judging legibility from the RStudio plot pane.

```{r}
#| eval: false
ggsave("forest_plot.pdf", width = 7, height = 5, dpi = 300)
```

## A Complete, Annotated Example

Pulling several of these refinements together, here is a more complete version of the `ggplot2` forest plot that you could reasonably drop into a report:

```{r}
plot_df <- studies %>%
  arrange(or) %>%
  mutate(study = factor(study, levels = study))

ggplot(plot_df, aes(x = or, y = study)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2, color = "grey30", linewidth = 0.6) +
  geom_point(aes(size = weight), shape = 15, color = "steelblue4") +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4), labels = label_number(accuracy = 0.01)) +
  scale_size_continuous(range = c(2, 5.5), guide = "none") +
  coord_cartesian(xlim = c(0.3, 3.5)) +
  labs(
    x = "Odds Ratio (95% CI, log scale)",
    y = NULL,
    title = "Effect of Drug X on 30-Day Hospital Readmission",
    subtitle = "Eight randomized controlled trials; point size reflects study weight",
    caption = "Dashed line indicates no effect (OR = 1). Values < 1 favor treatment."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0, color = "grey40", size = 9)
  )
```

This version sorts by effect size, uses a log-scaled and range-limited x-axis, scales point size by weight, strips out unnecessary gridlines, and adds a title, subtitle, and caption that together tell the reader everything they need without requiring them to already know forest plot conventions.

## Common Pitfalls

A few mistakes come up often enough to call out explicitly:

**Plotting ratio measures on a linear scale.** This is by far the most common error. If your effect measure is an odds ratio, risk ratio, or hazard ratio, the x-axis must be logarithmic, or the visual distances between points will misrepresent the underlying statistics — an OR of 2 and an OR of 0.5 represent equal and opposite effects, and they should look that way on the chart.

**Mismatched confidence level conventions.** Make sure the CI you are plotting (usually 95%, but sometimes 90% or 99% depending on the field) is clearly labeled, since a forest plot without a stated confidence level is ambiguous and can be misread as a different kind of interval (e.g. a prediction interval, which is a different and typically much wider quantity in meta-analysis).

**Confusing weights with sample size.** A study's weight in a meta-analysis is a function of its *variance*, not just its raw n — a small, extremely well-controlled study can outweigh a larger but noisier one. If you scale point size by `n` when your underlying model actually weighted by inverse variance, the visual and the statistics will disagree, which undermines trust in the figure.

**Forgetting to sort or group meaningfully.** As discussed above, unordered or arbitrarily-ordered rows make it much harder for a reader to spot patterns; even if you don't sort by effect size, some deliberate ordering (chronological, by subgroup, by study quality) is almost always better than whatever order the data happened to be entered in.

## Wrapping Up

Forest plots reward a bit of upfront thinking about what you're actually trying to communicate. If you are running a real meta-analysis, `metafor::forest()` will save you time and guarantee your plot and your statistics never drift out of sync — start there. If you need the classic clinical-report table-plus-chart layout and already have your summary numbers in hand, `forestplot` gets you there fast. And if you want full control, or your forest plot needs to sit alongside other `ggplot2` visuals in a broader report, building it with `geom_pointrange()` or `geom_errorbarh()` plus `geom_point()` gives you a flexible foundation you can extend indefinitely.

Whichever route you take, the core anatomy stays the same: one row per unit of comparison, a point estimate, a confidence interval, and a reference line that tells the reader what "no effect" looks like. Get those four things right, put the ratio measures on a log scale, sort with intent, and you'll have a forest plot that communicates as clearly as the best ones you've seen in print.

## Further Reading

- Viechtbauer, W. (2010). *Conducting Meta-Analyses in R with the metafor Package.* Journal of Statistical Software.
- Gordon, M. & Lumley, T. — documentation for the `forestplot` package on CRAN.
- Cochrane Handbook for Systematic Reviews of Interventions, Chapter 10: *Analysing data and undertaking meta-analyses.*
- Wickham, H. — *ggplot2: Elegant Graphics for Data Analysis*, for the underlying grammar-of-graphics concepts used throughout the `ggplot2` sections above.
