---
title: "Cutting Through Data: A Practical Guide to dplyr::slice() in R"
author: "Social Capital Research Institute"
format:
  html:
    theme: cosmo
    toc: true
    css: styles.css
execute:
  echo: true
  warning: false
  message: false
---

![](https://images.unsplash.com/photo-1647736070366-09ca6936f727?q=80&w=1600&auto=format&fit=crop){fig-align="center"}

## Introduction

There’s a quiet moment in data analysis when everything feels too big—too many rows, too much noise. You don’t want everything. You want something precise.

That’s where `slice()` from **dplyr** comes in. It gives you direct control over rows by position.

---

## Video Tutorial

{{< video https://www.youtube.com/watch?v=S0wPf9BI520 >}}

---

## Setup

```{r}
library(dplyr)

df <- tibble(
  id = 1:10,
  name = c("A","B","C","D","E","F","G","H","I","J"),
  score = c(50, 60, 55, 70, 65, 80, 75, 90, 85, 95)
)

df
```

---

## Basic Usage

```{r}
# select first row
df %>% slice(1)

# select specific rows
df %>% slice(2, 5, 7)

# select a range
df %>% slice(1:3)
```

---

## Removing Rows

```{r}
# remove first row
df %>% slice(-1)

# remove first three rows
df %>% slice(-(1:3))
```

---

## Grouped Data

```{r}
df %>%
  mutate(group = ifelse(id <= 5, "A", "B")) %>%
  group_by(group) %>%
  slice(1)
```

---

## Helper Functions

```{r}
# first n rows
df %>% slice_head(n = 3)

# last n rows
df %>% slice_tail(n = 3)

# highest values
df %>% slice_max(order_by = score, n = 3)

# lowest values
df %>% slice_min(order_by = score, n = 3)

# random sample
df %>% slice_sample(n = 3)
```

---

## Real Use Case

```{r}
df %>%
  mutate(group = ifelse(id <= 5, "A", "B")) %>%
  group_by(group) %>%
  slice_max(order_by = score, n = 2)
```

---

## Key Takeaway

- `filter()` → selects rows by condition  
- `slice()` → selects rows by position  

That difference is small—but powerful.

---

## Closing

`slice()` is simple, but once you start using it, it becomes one of those tools you reach for constantly. It gives you control over the structure of your data without overcomplicating things.